\pkg{sparseHessianFD}: An R Package for Estimating Sparse Hessian Matrices

### 2016-03-14 nobuild/jss/sparseHessianFD.tex
ready for JSS submission

\documentclass[article]{jss}

@@

 \newcolumntype{d}[1]{D{.}{.}{#1}}
\newcommand{\func}[1]{\code{#1}}
\newcommand{\class}[1]{\textsl{#1}}
\newcommand{\funcarg}[1]{\code{#1}}
\newcommand{\variable}[1]{\code{#1}}
\newcommand{\method}[1]{\func{#1}}

@@

\newenvironment{knitrout}{\begin{figure}}{\end{figure}}

\begin{document}

<<setup1, echo=FALSE, cache=FALSE>>=
suppressPackageStartupMessages(require(Matrix))
knitr::render_sweave()
knitr::opts_chunk$set(prompt=TRUE, cache=TRUE)
options(replace.assign=TRUE, width=90,prompt="R> ")
@

@@

<<setup2, echo=FALSE>>=
N <- 5
k <- 2
nv1 <- (N+1)*k
nels1 <- nv1^2
nnz1 <- (N+1)*k^2 + 2*N*k^2
nnz1LT <- (N+1)*k*(k+1)/2 + N*k^2
Q <- 1000
nv2 <- (Q+1)*k
nels2 <- nv2^2
nnz2 <- (Q+1)*k^2 + 2*Q*k^2
nnz2LT <- (Q+1)*k*(k+1)/2 + Q*k^2
options(scipen=999)
@

The sparsity pattern depends on how the variables are
ordered within the vector. One such ordering is to group all of the
coefficients for each unit together.
\begin{align}
\beta_{11},\dotsc,\beta_{1k},\beta_{21},\dotsc,\beta_{2k},~\dotsc~,\beta_{N1},\dotsc,\beta_{Nk},\mu_1,\dotsc,\mu_k
\end{align}
In this case, the Hessian has a ``block-arrow'' structure.  For example,
if $N=\Sexpr{N}$ and $k=\Sexpr{k}$, then there are \Sexpr{nv1} total
variables, and the Hessian will have the pattern in Figure~\ref{fig:blockarrow}.

Another possibility is to group coefficients for each covariate
together.
\begin{align}
\beta_{11},\dotsc,\beta_{N1},\beta_{12},\dotsc,\beta_{N2},~\dotsc~,\beta_{1k},\dotsc,\beta_{Nk},\mu_1,\dotsc,\mu_k
\end{align}
Now the Hessian has an "banded" sparsity pattern, as in Figure~\ref{fig:banded}.
\begin{figure}[tbp]
  \begin{subfigure}[b]{.5\textwidth}
<<blockarrow, echo=FALSE>>=
Mat <- as(kronecker(diag(N),matrix(1,k,k)),"sparseMatrix")
Mat <- rBind(Mat, Matrix(1,k,N*k))
Mat <- cBind(Mat, Matrix(1, k*(N+1), k))
printSpMatrix(as(Mat,"nMatrix"))
@
\caption{A ``block-arrow'' sparsity pattern.}\label{fig:blockarrow}
\end{subfigure}
\begin{subfigure}[b]{.5\textwidth}
<<banded, echo=FALSE>>=
Mat <- kronecker(Matrix(1,k,k), diag(N))
Mat <- rBind(Mat, Matrix(1,k,N*k))
Mat <- cBind(Mat, Matrix(1, k*(N+1), k))
printSpMatrix(as(Mat,"nMatrix"))
@
\caption{A ``banded'' sparsity pattern.}\label{fig:banded}
\end{subfigure}
\caption{Two examples of sparsity patterns for a hierarchical model.}\label{fig:patterns}
\end{figure}

In both cases, the number of non-zeros is the same.   There are
\Sexpr{nels1} elements in this symmetric matrix, but only \Sexpr{nnz1} are
non-zero, and only \Sexpr{nnz1LT} values are unique.  Although the reduction in
RAM from using a sparse matrix structure for the Hessian may be
modest, consider what would happen if $N=\Sexpr{scales::comma(Q)}$ instead.  In that case,
there are \Sexpr{scales::comma(nv2)} variables in the problem, and more than $\Sexpr{scales::comma(floor(nels2/10^6))}$ million
elements in the Hessian.  However, only $\Sexpr{scales::comma(nnz2)}$ of those elements are
non-zero.  If we work with only the lower triangle of the Hessian we only need to work with
only \Sexpr{scales::comma(nnz2LT)} values.

@@

\begin{minipage}{\columnwidth}
<<>>=
library("sparseHessianFD")
bd <- kronecker(diag(3), matrix(TRUE,2,2))
Mat <- as(bd, "nMatrix")
printSpMatrix(tril(Mat))
mc <- Matrix.to.Coord(tril(Mat))
mc
@
\end{minipage}

@@

<<>>=
pattern <- sparseMatrix(i=mc$rows, j=mc$cols)
printSpMatrix(pattern)
@

@@

To create a \class{sparseHessianFD} object, just call
\func{sparseHessianFD}.  Applying the default values for the optional
arguments,
<<eval=FALSE>>=
obj <- sparseHessianFD(x, fn, gr, rows, cols, ...)
@
where \funcarg{...} represents all other named arguments that are
passed to \funcarg{fn} and \funcarg{gr}.

@@

<<eval=FALSE>>=
f <- obj$fn(x)          ## returns numeric
df <- obj$gr(x)         ## returns numeric vector
hess <- obj$hessian(x)  ## returns dgCMatrix
fngr <- obj$fngr(x)     ## returns list
fngrhs <- obj$fngrhs(x) ## returns list
@

@@

<<binaryInit>>=
set.seed(123)
data("binary")
str(binary)
N <- length(binary[["Y"]])
k <- NROW(binary[["X"]])
T <- binary[["T"]]
nvars <- as.integer(N*k + k)
priors <- list(inv.Sigma = rWishart(1,k+5,diag(k))[,,1],
               inv.Omega = diag(k))
@

This dataset represents the simulated choices for $N= \Sexpr{N}$ customers
over $T= \Sexpr{T}$ purchase opportunties, where the probability of purchase
is influenced by $k= \Sexpr{k}$ covariates.

@@

<<trueValues>>=
P <- rnorm(nvars)
order.row <- FALSE
true.f <- binary.f(P, binary, priors, order.row=order.row)
true.grad <- binary.grad(P, binary, priors, order.row=order.row)
true.hess <- binary.hess(P, binary, priors, order.row=order.row)
@

@@

<<binaryRowsCols>>=
pattern <- Matrix.to.Coord(tril(true.hess))
str(pattern)
@

@@

<<usingSparseHessianFD>>=
obj <- sparseHessianFD(P, fn=binary.f, gr=binary.grad,
       rows=pattern[["rows"]], cols=pattern[["cols"]],
       data=binary, priors=priors, order.row=order.row)
f <- obj$fn(P)
all.equal(f, true.f)
gr <- obj$gr(P)
all.equal(gr, true.grad)
hs <- obj$hessian(P)
all.equal(hs, true.hess)
@

>>>>

\documentclass[article]{jss}\usepackage[]{graphicx}\usepackage[]{color}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

\usepackage{Sweave}

@@

   \newcolumntype{d}[1]{D{.}{.}{#1}}
%%\newcommand{\pkg}[1]{\emph{#1}}
%%\newcommand{\proglang}[1]{\textsf{#1}}
%%\newcommand{\filename}[1]{\textit{#1}}
%%\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\func}[1]{\code{#1}}
\newcommand{\class}[1]{\textsl{#1}}
\newcommand{\funcarg}[1]{\code{#1}}
\newcommand{\variable}[1]{\code{#1}}
\newcommand{\method}[1]{\func{#1}}

@@

\newenvironment{knitrout}{\begin{figure}}{\end{figure}}
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}

@@

The sparsity pattern depends on how the variables are
ordered within the vector. One such ordering is to group all of the
coefficients for each unit together.
\begin{align}
\beta_{11},\dotsc,\beta_{1k},\beta_{21},\dotsc,\beta_{2k},~\dotsc~,\beta_{N1},\dotsc,\beta_{Nk},\mu_1,\dotsc,\mu_k
\end{align}
In this case, the Hessian has a ``block-arrow'' structure.  For example,
if $N=5$ and $k=2$, then there are 12 total
variables, and the Hessian will have the pattern in Figure~\ref{fig:blockarrow}.

Another possibility is to group coefficients for each covariate
together.
\begin{align}
\beta_{11},\dotsc,\beta_{N1},\beta_{12},\dotsc,\beta_{N2},~\dotsc~,\beta_{1k},\dotsc,\beta_{Nk},\mu_1,\dotsc,\mu_k
\end{align}
Now the Hessian has an "banded" sparsity pattern, as in Figure~\ref{fig:banded}.
\begin{figure}[tbp]
  \begin{subfigure}[b]{.5\textwidth}
\begin{Schunk}
\begin{Soutput}

 [1,] | | . . . . . . . . | |
 [2,] | | . . . . . . . . | |
 [3,] . . | | . . . . . . | |
 [4,] . . | | . . . . . . | |
 [5,] . . . . | | . . . . | |
 [6,] . . . . | | . . . . | |
 [7,] . . . . . . | | . . | |
 [8,] . . . . . . | | . . | |
 [9,] . . . . . . . . | | | |
[10,] . . . . . . . . | | | |
[11,] | | | | | | | | | | | |
[12,] | | | | | | | | | | | |
\end{Soutput}
\end{Schunk}
\caption{A ``block-arrow'' sparsity pattern.}\label{fig:blockarrow}
\end{subfigure}
\begin{subfigure}[b]{.5\textwidth}
\begin{Schunk}
\begin{Soutput}

 [1,] | . . . . | . . . . | |
 [2,] . | . . . . | . . . | |
 [3,] . . | . . . . | . . | |
 [4,] . . . | . . . . | . | |
 [5,] . . . . | . . . . | | |
 [6,] | . . . . | . . . . | |
 [7,] . | . . . . | . . . | |
 [8,] . . | . . . . | . . | |
 [9,] . . . | . . . . | . | |
[10,] . . . . | . . . . | | |
[11,] | | | | | | | | | | | |
[12,] | | | | | | | | | | | |
\end{Soutput}
\end{Schunk}
\caption{A ``banded'' sparsity pattern.}\label{fig:banded}
\end{subfigure}
\caption{Two examples of sparsity patterns for a hierarchical model.}\label{fig:patterns}
\end{figure}

In both cases, the number of non-zeros is the same.   There are
144 elements in this symmetric matrix, but only 64 are
non-zero, and only 38 values are unique.  Although the reduction in
RAM from using a sparse matrix structure for the Hessian may be
modest, consider what would happen if $N=1,000$ instead.  In that case,
there are 2,002 variables in the problem, and more than $4$ million
elements in the Hessian.  However, only $12,004$ of those elements are
non-zero.  If we work with only the lower triangle of the Hessian we only need to work with
only 7,003 values.

@@

\begin{minipage}{\columnwidth}
\begin{Schunk}
\begin{Sinput}
R> library("sparseHessianFD")
R> bd <- kronecker(diag(3), matrix(TRUE,2,2))
R> Mat <- as(bd, "nMatrix")
R> printSpMatrix(tril(Mat))
\end{Sinput}
\begin{Soutput}

[1,] | . . . . .
[2,] | | . . . .
[3,] . . | . . .
[4,] . . | | . .
[5,] . . . . | .
[6,] . . . . | |
\end{Soutput}
\begin{Sinput}
R> mc <- Matrix.to.Coord(tril(Mat))
R> mc
\end{Sinput}
\begin{Soutput}
$rows
[1] 1 2 2 3 4 4 5 6 6

$cols
[1] 1 1 2 3 3 4 5 5 6
\end{Soutput}
\end{Schunk}
\end{minipage}

@@

\begin{Schunk}
\begin{Sinput}
R> pattern <- sparseMatrix(i=mc$rows, j=mc$cols)
R> printSpMatrix(pattern)
\end{Sinput}
\begin{Soutput}

[1,] | . . . . .
[2,] | | . . . .
[3,] . . | . . .
[4,] . . | | . .
[5,] . . . . | .
[6,] . . . . | |
\end{Soutput}
\end{Schunk}

@@

To create a \class{sparseHessianFD} object, just call
\func{sparseHessianFD}.  Applying the default values for the optional
arguments,
\begin{Schunk}
\begin{Sinput}
R> obj <- sparseHessianFD(x, fn, gr, rows, cols, ...)
\end{Sinput}
\end{Schunk}
where \funcarg{...} represents all other named arguments that are
passed to \funcarg{fn} and \funcarg{gr}.

@@

\begin{Schunk}
\begin{Sinput}
R> f <- obj$fn(x)          ## returns numeric
R> df <- obj$gr(x)         ## returns numeric vector
R> hess <- obj$hessian(x)  ## returns dgCMatrix
R> fngr <- obj$fngr(x)     ## returns list
R> fngrhs <- obj$fngrhs(x) ## returns list
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> set.seed(123)
R> data("binary")
R> str(binary)
\end{Sinput}
\begin{Soutput}
List of 3
 $ Y: int [1:50] 13 1 18 18 19 6 16 6 5 8 ...
 $ X: num [1:4, 1:50] -0.07926 -0.23018 1.55871 0.00997 0.01828 ...
 $ T: num 20
\end{Soutput}
\begin{Sinput}
R> N <- length(binary[["Y"]])
R> k <- NROW(binary[["X"]])
R> T <- binary[["T"]]
R> nvars <- as.integer(N*k + k)
R> priors <- list(inv.Sigma = rWishart(1,k+5,diag(k))[,,1],
+                inv.Omega = diag(k))
\end{Sinput}
\end{Schunk}

This dataset represents the simulated choices for $N= 50$ customers
over $T= 20$ purchase opportunties, where the probability of purchase
is influenced by $k= 4$ covariates.

@@

\begin{Schunk}
\begin{Sinput}
R> P <- rnorm(nvars)
R> order.row <- FALSE
R> true.f <- binary.f(P, binary, priors, order.row=order.row)
R> true.grad <- binary.grad(P, binary, priors, order.row=order.row)
R> true.hess <- binary.hess(P, binary, priors, order.row=order.row)
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> pattern <- Matrix.to.Coord(tril(true.hess))
R> str(pattern)
\end{Sinput}
\begin{Soutput}
List of 2
 $ rows: int [1:1310] 1 2 3 4 201 202 203 204 2 3 ...
 $ cols: int [1:1310] 1 1 1 1 1 1 1 1 2 2 ...
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> obj <- sparseHessianFD(P, fn=binary.f, gr=binary.grad,
+        rows=pattern[["rows"]], cols=pattern[["cols"]],
+        data=binary, priors=priors, order.row=order.row)
R> f <- obj$fn(P)
R> all.equal(f, true.f)
\end{Sinput}
\begin{Soutput}
[1] TRUE
\end{Soutput}
\begin{Sinput}
R> gr <- obj$gr(P)
R> all.equal(gr, true.grad)
\end{Sinput}
\begin{Soutput}
[1] TRUE
\end{Soutput}
\begin{Sinput}
R> hs <- obj$hessian(P)
R> all.equal(hs, true.hess)
\end{Sinput}
\begin{Soutput}
[1] TRUE
\end{Soutput}
\end{Schunk}
